# Partition values (quantiles)

Common partition values are:

1.  **Percentiles**
2.  **Quartiles**
3.  **Deciles**

## Percentiles

Percentiles are 99 values which divide a data set into approximately 100 equal parts. There are 99 percentiles, normally denoted as:

$$P_{1}, P_{2}, \dots, P_{25}, \dots, P_{50}, \dots, P_{75}, \dots, P_{99}$$

### A) Formula for Computing $j^{\text{th}}$ Percentile from Raw / Ungrouped Data

$$P_{j} = j^{\text{th}}\text{ Percentile} = \left[ j \times \frac{(n+1)}{100} \right]^{\text{th}}\text{ value}$$

::: {#exm-1}
Ten-count tests made on a certain yarn have shown the following results:\
`22.8, 23.2, 22.9, 22.6, 23.4, 23.0, 23.1, 23.0, 22.9, 23.0`

Compute $P_{25}$ and $P_{50}$.
:::

#### Solution

Here, $n = 10$.

**Arranged data:**\
`22.6, 22.8, 22.9, 22.9, 23.0, 23.0, 23.0, 23.1, 23.2, 23.4`

**Computing** $P_{25}$:

$$
\begin{aligned}
P_{25} &= \left[ 25 \times \frac{(10+1)}{100} \right]^{\text{th}}\text{ value} = 2.75^{\text{th}}\text{ value} \\
&= 2^{\text{nd}}\text{ value} + 0.75 \times (3^{\text{rd}}\text{ value} - 2^{\text{nd}}\text{ value}) \\
&= 22.8 + 0.75 \times (22.9 - 22.8) = 22.875
\end{aligned}
$$

**Computing** $P_{50}$:

$$
\begin{aligned}
P_{50} &= \left[ 50 \times \frac{(10+1)}{100} \right]^{\text{th}}\text{ value} = 5.5^{\text{th}}\text{ value} \\
&= 5^{\text{th}}\text{ value} + 0.5 \times (6^{\text{th}}\text{ value} - 5^{\text{th}}\text{ value}) \\
&= 23.0 + 0.5 \times (23.0 - 23.0) = 23.0
\end{aligned}
$$

------------------------------------------------------------------------

### B) Formula for Computing $j^{\text{th}}$ Percentile from Grouped Data / Frequency Distribution

$$P_{j} = L_{0} + \frac{c}{f} \left( \frac{N}{100} \times j - \text{p.cf} \right)$$

Where:

- $L_{0} = \text{Lower limit of the } j^{\text{th}} \text{ percentile class}$
- $c = \text{Class interval of the } j^{\text{th}} \text{ percentile class}$
- $f = \text{Frequency of the } j^{\text{th}} \text{ percentile class}$
- $\text{p.cf} = \text{Cumulative frequency of the class preceding the percentile class}$

::: {#exm-2}
The following data are related to the linear density of yarn:

| Linear density | 13.00–13.25 | 13.25–13.50 | 13.50–13.75 | 13.75–14.00 | 14.00–14.25 | 14.25–14.50 | 14.50–14.75 |
|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
| **No. of tests** | 8 | 12 | 20 | 25 | 22 | 10 | 4 |

Compute and explain $P_{50}$ and $P_{75}$.
:::

#### Solution

First, we compute the cumulative frequency ($cf$):

| Linear density  | Frequency ($f$) | Cumulative Frequency ($cf$) |
|:----------------|:---------------:|:---------------------------:|
| 13.00–13.25     |        8        |              8              |
| 13.25–13.50     |       12        |             20              |
| 13.50–13.75     |       20        |             40              |
| **13.75–14.00** |     **25**      |           **65**            |
| 14.00–14.25     |       22        |             87              |
| 14.25–14.50     |       10        |             97              |
| 14.50–14.75     |        4        |             101             |
| **Total**       |    $N = 101$    |                             |

**Finding** $P_{50}$:

To locate the $50^{\text{th}}$ percentile class, compute:

$$\frac{N}{100} \times 50 = \frac{101}{100} \times 50 = 50.5$$

The cumulative frequency just greater than $50.5$ is $65$.\
So, **13.75–14.00** is the $50^{\text{th}}$ percentile class.

$$
\begin{aligned}
P_{50} &= L_{0} + \frac{c}{f} \left( \frac{N}{100} \times 50 - \text{p.cf} \right) \\
&= 13.75 + \frac{0.25}{25} \times (50.5 - 40) \\
&= 13.855
\end{aligned}
$$

**Interpretation:**\
This means 50% of the 101 observations (i.e., $50\% \times 101 = 50.5 \approx 51$ observations) have a linear density lying at or below $13.855$.

*(Try computing* $P_{75}$ yourself as practice.)

------------------------------------------------------------------------

## Quartiles

Quartiles are 3 values that divide a data set into approximately 4 equal parts:

1.  $Q_{1}$ ($1^{\text{st}}$ quartile / lower quartile)
2.  $Q_{2}$ ($2^{\text{nd}}$ quartile / median)
3.  $Q_{3}$ ($3^{\text{rd}}$ quartile / upper quartile)

Note that:

- $Q_{1} = P_{25}$
- $Q_{2} = P_{50} = \text{Median}$
- $Q_{3} = P_{75}$

### Formula for Raw Data

$$Q_{j} = \left( j \times \frac{n+1}{4} \right)^{\text{th}}\text{ value}, \quad \text{for } j = 1, 2, 3$$

::: {#exm-3}
Find $Q_{1}$, $Q_{2}$, and $Q_{3}$ for the following data representing CPU time of 11 jobs (in seconds):\
`59, 139, 46, 37, 42, 30, 55, 56, 36, 82, 29`
:::

#### Solution

Here, $n = 11$.

**Sorted data:**\
`29, 30, 36, 37, 42, 46, 55, 56, 59, 82, 139`

$$
\begin{aligned}
Q_{1} = P_{25} &= \left( 1 \times \frac{11+1}{4} \right)^{\text{th}}\text{ value} = 3^{\text{rd}}\text{ value} = 36 \\
Q_{2} = P_{50} &= \left( 2 \times \frac{11+1}{4} \right)^{\text{th}}\text{ value} = 6^{\text{th}}\text{ value} = 46 \\
Q_{3} = P_{75} &= \left( 3 \times \frac{11+1}{4} \right)^{\text{th}}\text{ value} = 9^{\text{th}}\text{ value} = 59
\end{aligned}
$$

### Interquartile Range (IQR)

The **Interquartile Range** is the difference between the third quartile ($Q_{3}$) and the first quartile ($Q_{1}$):

$$\text{IQR} = Q_{3} - Q_{1}$$

------------------------------------------------------------------------

::: {#exr-interpretation}
### Question on Interpretation

In a class of 40 students, mid-term test marks were collected and $P_{20}$ was found to be 23.

1.  What does $P_{20} = 23$ mean?
2.  Approximately how many students received a score of 23 or less?

**Answer:**\
$P_{20} = 23$ means that 20% of the students scored 23 or fewer marks.\
Number of students scoring 23 or less = $40 \times 20\% = 8$ students.
:::

------------------------------------------------------------------------

## Box-Plots and the Five-Number Summary

**Five-Number Summary**

The **minimum**, **first quartile (**$Q_1$), **median (**$Q_2$), **third quartile (**$Q_3$), and **maximum**, taken together, are referred to as the **five-number summary** of a data set.

### Box-and-Whisker Plot

A **box-plot** (or box-and-whisker plot) is a chart used in exploratory data analysis to visually display the distribution of numerical data and its skewness through the five-number summary.

```{r}
#| label: fig-simple-boxplot
#| fig-cap: "Simple Box-plot of CPU Time (in seconds)"
#| echo: false
#| fig-width: 6
#| fig-height: 2.5

x <- c(59, 46, 37, 42, 30, 55, 56, 36, 82, 29, 139)
par(mar = c(3, 1, 2, 1))
boxplot(x, horizontal = TRUE, main = "Simple box-plot", col = "gray80", xlab = "CPU Time (seconds)")
```

### Significance of a Box-Plot

- Provides a quick visual summary of data distribution.
- Shows the central tendency from the line inside the box (Median).
- Displays total spread (Minimum to Maximum) as well as middle 50% spread ($\text{IQR} = Q_3 - Q_1$).
- Gives clear indications about the skewness/shape of the distribution.

------------------------------------------------------------------------

### Parallel / Comparative Box-Plots

Box-plots are frequently used to compare multiple groups or populations side-by-side using the same scale.

### Example: Tensile Strength of Paper

*(Adapted from D.C. Montgomery, Applied Statistics and Probability for Engineers)*

A paper manufacturer wants to improve product tensile strength. Engineering tests four levels of hardwood concentration in paper pulp: 5%, 10%, 15%, and 20%. Six test specimens are prepared at each concentration level.

#### Data Table

| Hardwood Conc. (%) | Obs 1 | Obs 2 | Obs 3 | Obs 4 | Obs 5 | Obs 6 | Totals  | Averages  |
|:------:|:------:|:------:|:------:|:------:|:------:|:------:|:------:|:------:|
|       **5%**       |   7   |   8   |  15   |  11   |   9   |  10   |   60    |   10.00   |
|      **10%**       |  12   |  17   |  13   |  18   |  19   |  15   |   94    |   15.67   |
|      **15%**       |  14   |  18   |  19   |  17   |  16   |  18   |   102   |   17.00   |
|      **20%**       |  19   |  25   |  22   |  23   |  18   |  20   |   127   |   21.17   |
|    **Overall**     |       |       |       |       |       |       | **383** | **15.96** |

```{r}
#| label: fig-comparative-boxplot
#| fig-cap: "Box plots of hardwood concentration data (Montgomery, Table 13.1)"
#| echo: false
#| fig-width: 6
#| fig-height: 4.5

tensile_data <- data.frame(
  strength = c(7,8,15,11,9,10, 12,17,13,18,19,15, 14,18,19,17,16,18, 19,25,22,23,18,20),
  concentration = factor(rep(c("5%", "10%", "15%", "20%"), each = 6), levels = c("5%", "10%", "15%", "20%"))
)

boxplot(strength ~ concentration, data = tensile_data,
        xlab = "Hardwood concentration (%)",
        ylab = "Tensile strength (psi)",
        col = "lightblue",
        main = "Tensile Strength by Hardwood Concentration")
```

#### Interpretation

- **Effect on Strength:** Increasing hardwood concentration consistently increases mean and median tensile strength.
- **Symmetry & Variability:** Distribution across levels is relatively symmetric, and dispersion/variability does not drastically change across treatment groups.

------------------------------------------------------------------------

## Exercises

1.  **Daily Wages:** Find the $3^{\text{rd}}$ quartile and $60^{\text{th}}$ percentile for the following distribution of daily wages of temporary workers, and comment on the results:

| Daily wages (Rs)   | 40–49 | 50–59 | 60–69 | 70–79 | 80–89 |
|:-------------------|:-----:|:-----:|:-----:|:-----:|:-----:|
| **No. of workers** |  15   |  20   |  30   |  45   |  25   |

------------------------------------------------------------------------

2.  **Fabric Strength:** Calculate the $3^{\text{rd}}$ decile ($D_3$), $2^{\text{nd}}$ percentile ($P_2$), and $70^{\text{th}}$ percentile ($P_{70}$):

| Fabric strength    | 45–49 | 50–54 | 55–59 | 60–64 | 65–69 | 70–74 |
|:-------------------|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
| **No. of samples** |   5   |  10   |  15   |  20   |  10   |   5   |

------------------------------------------------------------------------

3.  **Coefficient of Variation (CV%):** Calculate $Q_1$ and $P_{70}$ for the following data and comment on them:

| CV%              | 3.0–3.5 | 3.5–4.0 | 4.0–4.5 | 4.5–5.0 | 5.0–5.5 |
|:-----------------|:-------:|:-------:|:-------:|:-------:|:-------:|
| **No. of tests** |   25    |    4    |   10    |   16    |   10    |

------------------------------------------------------------------------

4.  **Goods Weight:** Compute the Median and $38^{\text{th}}$ percentile ($P_{38}$) for the following data:

| Weight (kg)      | 10–19 | 20–29 | 30–39 | 40–49 | 50–59 | 60–69 |
|:-----------------|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
| **No. of goods** |   6   |  10   |  16   |  14   |   8   |   4   |

------------------------------------------------------------------------

5.  **Yarn Unevenness (U%):** Compute the $1^{\text{st}}$ quartile ($Q_1$), $7^{\text{th}}$ decile ($D_7$), and $38^{\text{th}}$ percentile ($P_{38}$) for the following raw measurements:\
    `2.25, 4.25, 2.65, 3.85, 5.82, 3.42, 4.44, 2.86, 2.88, 4.46`

------------------------------------------------------------------------

6.  **Loom Comparison:** A textile company measures fabric strength (psi) across 2 randomly selected looms:

| Loom  | Measurements                   |
|:------|:-------------------------------|
| **A** | 99, 97, 97, 96, 97, 96, 92, 98 |
| **B** | 94, 95, 90, 92, 93, 94, 90, 92 |

i.  Construct a parallel/comparative box-plot comparing fabric strength for Loom A and Loom B.
ii. Based on the box-plot, which loom produces better fabric in terms of strength? Which loom exhibits lower variability?
